---
title: "Dice and dose-response relationships"
output: tint::tintPdf
latexfonts:
  - package: newtxmath
    options: 
      - cmintegrals
      - cmbraces
  - package: ebgaramond-maths
header-includes:
   - \usepackage{booktabs}
   - \usepackage{cancel}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      echo = FALSE,
                      cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

```{r, include=FALSE}
 library(tidyverse)
# library(scales) # for prettier axis labels
# library(simecol) # for running model 
theme_set(theme_classic())
library(ggpubr)
```

Dose-response relationships frequently show up in biological (and other) systems. Indeed, I bet you can think of at least four _different_ ways some _response_ changes with the dose of some _cause_. Let's write those down:

1.   
2.   
3.  
4.  

In fact, draw graphs of those relationships. Add labels (including units if you can) and put reasonable numbers on the two axes. 

```{r, fig.height=3}
p <- ggplot(tibble(x=0:1, y=0:1), aes(x,y)) + 
  geom_blank() + 
  theme(text = element_text(color = "white"), 
        axis.text=element_text(color="white"),aspect.ratio = 1
        )

ggarrange(p,p,p,p, labels = 1:4, nrow = 2, ncol=2)

```

It's good to keep this diversity in mind if only so that you don't make the mistake of assuming every dose-response relationship means the same thing.

# Dice as pathogens

To develop a more intuitive sense of at least one way we can get a dose-response relationship, let's roll some dice. I like thinking about pathogens, so let's imagine that we are the hosts and each die is a single pathogen. A pathogen/die only causes an infection if it comes up a "6", otherwise you fight it off. We are going to simulate exposures to different doses of pathogens/dice to see how the probability of infection changes with the dose.

Of course a single host is either infected or uninfected, so a single person's data would like a series of zeros and ones. We want to determine what the _probability_ of infection is at different doses, which we can estimate as the _proportion_ of hosts^[Students rolling dice.] that are infected^[Roll at least one 6.] at each dose.

```{r}
df <- data.frame(Dice = c(1, 2, 3, 4, 5, 7, 10, 15, 20, 30, 40),
             `Number of Hosts` = "  ",
             `Number Infected` = "  ",
             Proportion = " ") #1-(1-1/6)^Dice)

knitr::kable(df)
```

Then let's plot these data on two graphs, one with a linear x-axis and one with a logarithmic x-axis.

```{r}
df <- tibble(Dice = c(1, 2, 3, 4, 5, 7, 10, 15, 20, 30, 40),
             Prob = 1-(1-1/6)^Dice)

ggplot(df, aes(Dice, Prob)) + 
  # geom_point() + 
  geom_blank() + 
  scale_x_continuous("Dice rolled", 
                     breaks = df$Dice,
                     minor_breaks = NULL) + 
  scale_y_continuous("Probability of at least one 6", 
                     limits = c(0,1),
                     breaks = 0:10/10) + 
  theme_bw()

last_plot() + 
  scale_x_log10("Dice rolled", 
                breaks = df$Dice, 
                minor_breaks = NULL)

```

