---
title: "Dice and dose-response relationships: toxins"
author-meta: "Jesse Brunner"
date-meta: "`r Sys.Date()`"
format:
  pdf: default
  html: 
    self-contained: true
    grid:
      margin-width: 350px
format-links: false
execute:
  echo: false
toc: false
reference-location: margin
---


```{r, include=FALSE}
 library(tidyverse)
# library(scales) # for prettier axis labels
# library(simecol) # for running model 
theme_set(theme_classic())
library(ggpubr)
```

Dose-response relationships frequently show up in biological (and other) systems. Indeed, I bet you can think of at least four _different_ ways some _response_ changes with the dose of some _cause_. Let's write those down:

1.   
2.   
3.  
4.  

In fact, draw graphs of those relationships. Add labels (including units if you can) and put reasonable numbers on the two axes. 

```{r}
#| fig-column: page-right
#| fig-height: 3
#| fig-width: 9
#| fig-alt: "Four sets of blank axes to use to draw dose response relationships"
p <- ggplot(tibble(x=0:1, y=0:1), aes(x,y)) + 
  geom_blank() + 
  theme(text = element_blank(), 
        aspect.ratio = 1
        )

ggarrange(p,p,p,p, labels = 1:4, nrow = 1, ncol=4, spacing = 0.5)

```

It's good to keep this diversity in mind if only so that you don't make the mistake of assuming every dose-response relationship means the same thing.

# Dice determining susceptibility

To develop a more intuitive sense of at least one way we can get a dose-response relationship, let's roll some dice. 

Imagine that each die represents some component of the mechanism(s) that help you remove, inactivate, repair damage from, or otherwise tolerate some toxin X^[Imagine something specific if it helps, but it is not necessary for our little game.]. For each die/component you might have a low (1) or high (6) value, or anything in between. The sum of values across all of your dice/mechanisms determines your _tolerance_ of the toxin. That is, it determines how much exposure you can tolerate before you succumb. 

Here we go:

  1. Please roll 10 dice and add up the numbers. Write that down here: ______
  2. Divide that number by 10 and write it down here: _____ _<--This is your susceptibility_


```{r simulating the activity}
#| include: false
nstudents <- 50
ndice <- 10

# Simulate dice rolls
sum_dice <- numeric()
for(i in 1:nstudents){
  sum_dice[i] <- sum( sample(1:6, ndice, replace = T) )  
}

# calculate susceptibility
suscept <- sum_dice/ndice

stem(suscept)
hist(suscept, breaks = 7)

cumsum( table(suscept) )/nstudents

# now a dose-response design with these orders of magnitude
doses <- c(2, 2.5, 3, 3.5, 4, 4.5, 5)

# how many students would be infected at each dose?
infected <- numeric()

for(i in 1:length(doses)){
  infected[i] <- sum(suscept <= doses[i])
}

plot( infected/nstudents ~ doses, type = "b")
```


# A dose-response experiment

At this point, every person in class should have their own (hidden) level of susceptibility. _I_, the experimenter, do not know anyone's susceptibility to toxin X, nor do even know anything useful like the average susceptibility of the class or the LD~50~. Let's see how I can use a dose-response study to figure out some of this.

Imagine I exposed the class to a particular dose of toxin X. I could see how many individuals (or the fraction of individuals) that _responded_ to the exposure (i.e., died if that was our outcome of interest). That could then be plotted on the axes, below. The position along the x-axis is the dose and the position on the y-axis is the proportion of the class that responded. 


```{r}
df <- tibble(Log10Dose = c(2, 2.5, 3, 3.5, 4, 4.5, 5),
             Dose = 10^Log10Dose,
             Prop = c(0, 0.05, 0.15, 0.4, 0.8, 0.9, 1))

ggplot(df, aes(Dose, Prop)) + 
  # geom_point() + 
  geom_blank() + 
  scale_x_log10("Exposure Dose", 
                     breaks = df$Dose,
                     minor_breaks = NULL, 
                label = scales::label_log()
                ) + 
  scale_y_continuous("Proportion of the class that respond", 
                     limits = c(0,1),
                     breaks = 0:10/10) + 
  theme_bw()
```

Let's try this. I will expose the class to the dose and you simply raise your hand if you would respond to this dose. You should respond only if the _exponent_ of the dose (i.e., the 2.5 in a dose 10^2.5^) is _higher_ than your tolerance (the number you wrote above).


|                Dose | 10^2^ | 10^2.5^ | 10^3^ | 10^3.5^ | 10^4^ | 10^4.5^ | 10^5^ |
|--------------------:|:-----:|:-------:|:-----:|:-------:|:-----:|:-------:|:-----:|
| Number of responses |       |         |       |         |       |         |       |


Now let's plot these data. 

Do they look like what you expected? What would happen if we repeated this "experiment"? Would we get the same or similar results? Why?

# Useful and not

As always, our little toy model is useful in some ways and not-so-useful in other ways. Think about what was useful for helpful about this activity and what is missing, inaccurate, misleading, or otherwise not so helpful.
