---
title: "Dice and dose-response relationships: infections"
author-meta: "Jesse Brunner"
date-meta: "`r Sys.Date()`"
format:
  pdf: default
  html: 
    self-contained: true
    grid:
      margin-width: 350px
format-links: false
execute:
  echo: false
toc: false
reference-location: margin
---


```{r, include=FALSE}
 library(tidyverse)
# library(scales) # for prettier axis labels
# library(simecol) # for running model 
theme_set(theme_classic())
library(ggpubr)
```

We have seen dose-response relationships in several contexts, from fertilizing with limiting nutrients increasing growth rates to increasing doses of toxins overwhelming an individual's capacity to tolerate it. Today we will think about dose-responses in pathogen or parasite infections^[The logic also holds for other cases where each challenge from something---a spark that might light a fire, a hit to the head that might cause a traumatic brain injury---has a small, independent chance of causing a catastrophe.]. 

# Dice as pathogens

To develop a more intuitive sense of at least one way we can get a dose-response relationship, let's roll some dice. I like thinking about pathogens, so let's imagine that we are the hosts and each die is a single pathogen. A pathogen/die only causes an infection if it comes up a "6", otherwise you fight it off. We are going to simulate exposures to different doses of pathogens/dice to see how the probability of infection changes with the dose.

Of course a single host is either infected or uninfected, so a single person's data would either be a zero or a one (or a series of zeros and ones with repeated exposures). We want to determine what the _probability_ of infection is at different doses, which we can estimate as the _proportion_ of hosts^[Students rolling dice.] that are infected^[Roll at least one 6.] at each dose.

We will also consider two other scenarios: one where _two_ sixes in a roll are needed to cause an infection and one where _four_ are needed. 

Before we start rolling our dice, I would like you to sketch out on the axes below your expectation for what these data are going to look like for each criterion for a successful infection. Be sure to think about what you will see for each of our criteria for success.


```{r}
#| include: false


# Actual data from class in 2024
real <- tibble(
  N = 38,
  Dice= c( 1, 2, 3, 4, 5, 7,10,15,20,30,40),
  One = c( 7, 9,11,15,19,22,24,35,38,38,38), 
  Two = c( 0, 0, 0, 2, 5, 8,22,29,37,36,36), 
  Four= c( 0, 0, 0, 0, 0, 1, 3,12,14,24,34)
) %>% 
  pivot_longer(c(One, Two, Four), names_to = "Criterion", values_to = "Infected")

# Theoretical values given the dice rules
theory <- tibble(
  Dice= c(1:5, 7,10,15,20,30,40),
  One = 1-pbinom(0, size=c(1:5, 7,10,15,20,30,40), prob=1/6),
  Two = 1-pbinom(1, size=c(1:5, 7,10,15,20,30,40), prob=1/6),
  Four = 1-pbinom(3, size=c(1:5, 7,10,15,20,30,40), prob=1/6)
) %>% 
  pivot_longer(-Dice, names_to = "Criterion", values_to = "Probability")


ggplot(real, aes(x=Dice, y=Infected/N, color = Criterion)) + 
  geom_point() + geom_line() +
  geom_line(data=theory, aes(y=Probability), linetype=2) + 
  scale_x_log10() + 
  scale_y_continuous("Proportion infected")
```

Number of students: ______

```{r}
df <- data.frame(Dice = c(1, 2, 3, 4, 5, 7, 10, 15, 20, 30, 40),
             One = "  ", Two = "  ", Four = " ") 

knitr::kable(df, col.names=c("Dice rolled", "Students with $\\geq$ 1", "Students with $\\geq$ 2", "Students with $\\geq$ 4"))
```

Then let's plot these data on the axes, below. (And yes, you will have to turn numbers of "infected" students into proportions of the class.)

```{r}
df <- tibble(Dice = c(1, 2, 3, 4, 5, 7, 10, 15, 20, 30, 40),
             Prob = 1-(1-1/6)^Dice)

ggplot(df, aes(Dice, Prob)) + 
  # geom_point() + 
  geom_blank() + 
  scale_x_continuous("Dice rolled", 
                     breaks = df$Dice,
                     minor_breaks = NULL) + 
  scale_y_continuous("Probability of at least one 6", 
                     limits = c(0,1),
                     breaks = 0:10/10) + 
  theme_bw()
```

Do your data look like what you expected? Why? What fits your expectation and what is a surprise?

\vspace{0.25cm}

Now plot the same data on the _logarithmic_ x-axis, below. Note that distance along the x-axis now reflects _proportional_ increases (e.g., doubling is always the same) rather than the increases in number. 
```{r}
#| message: false
last_plot() + 
  scale_x_log10("Dice rolled", 
                breaks = df$Dice, 
                minor_breaks = NULL)

```

Does this pattern fit your expectations better? Worse? 

# Useful and not

As always, our little toy model is useful in some ways and not-so-useful in other ways. Think about what was useful for helpful about this activity and what is missing, inaccurate, misleading, or otherwise not so helpful.
