---
title: "Drawing graphs of causal relationships"
output: tint::tintPdf
latexfonts:
  - package: newtxmath
    options: 
      - cmintegrals
      - cmbraces
  - package: ebgaramond-maths
header-includes:
   - \usepackage{booktabs}
   - \usepackage{cancel}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      echo = FALSE,
                      cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

```{r, include=FALSE}
# library(tidyverse)
# library(scales) # for prettier axis labels
# library(simecol) # for running model 
# theme_set(theme_minimal())
```

# Graphs and BHC

Often we see evidence presented in the form of a graph. Good graphs will tell a story of sorts, highlighting relationships or comparisons. _Drawing_ graphs will help you think about what how you think things work or what aspects of the relationship(s) you want to illustrate. We'll be doing a lot of graph drawing in this class! 

To get us started, I would like you and your group to come up with some _causal_ relationship that you find at least moderately interesting. Let me know if you need help with this, but once you have it, write it down here:


\vspace{2cm}

Then I would like you to draw graphs (or sets of graphs) that illustrate for each of the following "criteria" of Bradford Hill made up data^[Yes, you can make up data. It's a good thought exercise to think about what sort of data would be helpful for a particular argument.] that would _support_ this causal assertion. You can use what ever type of graph(s) you like. You might even try different types of each graph to see which is most compelling. Please label the axes as best you can and be precise, within the limits of your understanding of the system.

_Then_, please draw a second set of graphs, using identical forms and axes, showing what the data would look like if this relationship were _not_ causal.

## Strength of association or effect size
\vspace{4cm}

## Biological gradient (dose-effect)

\vspace{4cm}

## Consistency (reproducibility)

\vspace{5cm}

## Coherence
\vspace{5cm}

## Specificity
\vspace{5cm}

## Temporality
\vspace{5cm}
