---
title: "Parasites, commensals, & amensals, oh my!"
output: tint::tintPdf
latexfonts:
  - package: newtxmath
    options: 
      - cmintegrals
      - cmbraces
  - package: ebgaramond-maths
header-includes:
   - \usepackage{booktabs}
   - \usepackage{cancel}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      echo = FALSE,
                      cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

```{r, include=FALSE}
library(tidyverse)
library(scales) # for prettier axis labels
library(simecol) # for running model 
theme_set(theme_minimal())
```

# Graphing your interactions
We have all dealt with a variety of people in many different settings that have left us with good, neutral, and sour feelings. We've felt cheated. We've felt like we benefited from an interaction. Most interactions, I would guess, haven't left much of an impression on us at all. But I'm also confident that you haven't spent a lot of time _graphing_ those interactions. Well, today is your lucky day!

```{r}
df <- expand_grid(A = -1:1, B = -1:1) %>% 
  mutate(label = c("Competition", "amensalism", "pred/herb/parasit",
                   "amensalism", "neutralism", "commensalism",
                   "pred/herb/parasit", "commensalism", "mutualism"
         ))

ggplot(df, aes(A, B)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_text(aes(label = label), angle = 30, alpha = 1/3, size = 2) + 
  labs(y = "Effect on Player B", x = "Effect on Player A (you)") +
  coord_equal(ylim = c(-1.5, 1.5), xlim = c(-1.5, 1.5)) +
  theme_minimal() + 
  theme(axis.title = element_text(size = 5), 
        axis.text = element_text(size = 5))
  
```

I would like you (and your group) to **plot**^[Note: Each interaction need not be represented by a point, but could be a line, region, or whatever shape makes the most sense to you.] **these  interactions** on the axes above. If you and your group disagree about where an interaction goes, be sure to discuss _why_. Please focus on the good and bad for you as an individual and the other player, as opposed to societal good and bad.

* you & your credit card company
* you & a grocery store
* you & your teachers
* you & a cook or chef
* you as the babysitter for a child
* you & a sibling
* you & a parent
* you & the Pullman City government

Please then **plot your group's points or regions on the whiteboard** with everyone else's. If there is variability among groups, why might that be? Are the other groups simply wrong?

Now, think about how these interactions might change, and what factors or forces would change them. Do they all respond to the same factor(s)? Do they all change in the same direction?

Please come up with **three interspecific interactions** (i.e., between species) and plot them on the axes, above. 

Consider why these interspecific interactions might change. Are there patterns or common explanations for how they change? Do these explanations relate to concepts we've discussed in this class?
