---
title: "Some perspective on large numbers"
# author: "Jesse Brunner"
# date: "`r Sys.Date()`"
output: tint::tintPdf
latexfonts:
  - package: newtxmath
    options: 
      - cmintegrals
      - cmbraces
  - package: ebgaramond-maths
header-includes:
   - \usepackage{booktabs}
   - \usepackage{cancel}
   - \usepackage[version=4]{mhchem}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      echo = FALSE,
                      cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

```{r, include=FALSE}
library(tidyverse)
library(scales) # for prettier axis labels
# library(simecol) # for running model 
theme_set(theme_minimal())
```



Let's be honest: it can be hard to think about big numbers in any very intuitive sense. In an evolutionary sense, we were never exposed to and so never got good at thinking with such large numbers^[Not until the invention of mathematics beyond simply counting and science, anyway!]  After a certain number of zeros, everything gets compressed in our thinking to "big"! A million? A billion? A trillion? They're just much larger than I can count. Or are they?

Have you ever tried counting to a million? I'm guessing you have better things to do with your time, but using the power of Math~TM~ we could figure out how long it would take if you _did_ count to a million, and a billion, and so on. So let's figure this out. 

# Once per second

If we made the simplifying assumption that you count one number second then counting to a million is pretty straight-forward: it takes a million seconds! 

Is that not intuitive? Then let's try converting that to, say, hours^[Notice that all of the units cancel out.]. 

$$
1,000,000 \text{ numbers} \times \frac{1 \text{ second}}{ \text{ number}} \times \frac{\text{minute}}{60 \text{ seconds}}\times \frac{\text{hour}}{60 \text{ minutes}} = 277.7\bar{7} \text{ hours}
$$
Well that's helpful, right? No? Still not intuitive? Well let's convert it to days, then. 

$$
1,000,000 \text{ numbers} \times \frac{1 \text{ second}}{ \text{ number}} \times \frac{\text{minute}}{60 \text{ seconds}}\times \frac{\text{hour}}{60 \text{ minutes}} \times \frac{\text{day}}{24 \text{ hours}} =
$$
I'll let you do the calculation. I'm so nice, right?

# But I count faster!

I agree, you can probably count faster than one number second (up to a point anyway). So let's see how much you can reduce this time to count to a million.

`r newthought("Please time yourself counting")` from one to one-hundred. Then convert that into a value for $\text{seconds}/\text{number}$. Now redo the calculation for how long it would take you to count to a million if you were able to keep up this same pace. 
\vspace{2cm}

# Now a billion

Let's repeat these calculations for a _billion_. And then a _trillion_. 

\vspace{4cm}

_Now_ do you have some sense for how large these numbers are? And remember, moving from a million to a billion is an increase of a thousand. Moving from a million to a trillion is an increase of a thousand-thousand, or a million. In other words, a trillion is a million-million. It is a much, much, much bigger number.

# But why?

Why do we care? Because in science we often end up dealing with large numbers. And we need to be able to work with them. So, for instance, if I were to ask you if there are more humans on earth than (human) cells in the human body, what would you say?

The answer is that, as of this writing, there are an [estimated](https://www.census.gov/popclock/) 7,924,962,530 people on earth and an [estimated](https://pubmed.ncbi.nlm.nih.gov/23829164/) $37.2 \times 10^{12}$ cells in an average human body^[[This is a lovely little description](https://www.nationalgeographic.com/science/article/how-many-cells-are-in-your-body) of how they estimated this number... obviously no one wanted to _count_ them all! ].

How many virus particles might be in an infected person? For SARS-CoV-2 [one study](https://www.pnas.org/doi/full/10.1073/pnas.2024815118) estimated a person might carry between 1 billion and 100 billion virus particles (or virions). Which is a lot, but not nearly as many virus particles as cells (as is implied in the [very wrong headline of this news piece](https://www.news-medical.net/news/20201118/Around-10-SARS-CoV-2-infected-virions-per-cell-in-human-body-study-finds.aspx) on the study). Amazingly the authors then estimated the _mass_ of all of the SARS-CoV-2 virions in all of humans. Any guesses?^[No, I'm not telling you! Follow the links for the answer.] 

# A challenge

Let's think about this growing SARS-CoV-2 population. If we assume an infection starts with $N_0 = 1$ infectious virion and grows to a billion virions in the span of five days, what must its replication rate be? 

Let's assume continuous, constant replication (and ignore the immune system fighing back and all of the other complexities), so we can use the exponential model of growth:

$$
N(t) = N_0 e^{rt}
$$
Start by assigning the known values I've given you to the variables in this equation. (Hint: we know $t = 5$ days, right?) Then remember that these are _equations_ can can be manipulated any which way you like^[But preferably according to the rules of mathematics.]. You should be able to estimate $r$. Give it a shot!

\vspace{3cm}

Now try if we assume it grows to ten-billion virions. Or starts from 10 virions.

\vspace{3cm}

# A final challenge

Now let's consider the number of cells in a human newborn baby. I found an [estimate](https://bionumbers.hms.harvard.edu/bionumber.aspx?id=106413&ver=4)^[Apparently based on a pretty old paper, so_ caveat emptor_] of $1.25 \times 10^{12}$ cells. We also know that the newborn began from a single cell roughly nine months earlier. So... if we assume that the cells all divide in two, in discrete steps^[This is only true for a little while, but let's run with this assumption.], how many times does that one cell need to divide to produce the 1.25 trillion it is born with?

It may be helpful to recall this equation:

$$
N(t) = N_0 \times \lambda^t, 
$$
since we are assuming discrete time steps. Can you figure it out? Is it more or less than you expected?

